---
title: "Ice cover duration"
author: "Peter Menzies"
date: "8/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(ggExtra)
```

```{r}
ice_cover <- read_csv(here("data", "ntl33_v7.csv"), na = c("NA", "-999"))

weather <- read_csv(here("data", "ntl20_v6.csv"))
```

```{r}
ice_cover <-ice_cover %>% 
  group_by(lakeid) %>% 
  mutate(median =  median(ice_duration, na.rm = TRUE))

ice_plot <- ggplot(ice_cover, aes(x = lakeid, y = ice_duration)) +
  geom_jitter() +
  labs(x = "Lake", y = "Number of days frozen", title = "\n\nNumber of frozen days at three Madison Lake Area lakes")

ggMarginal(ice_plot, type = "boxplot", groupColour = TRUE, outlier.shape = NA)
```

### Find the mean ice cover duration in Madison Lake Area

```{r}
mean_ice_cover <- ice_cover %>% 
  group_by(year4) %>% 
  summarize(mean = mean(ice_duration, na.rm = TRUE)) %>% 
  filter(mean != "NaN")


mean_yearly_ice <- ggplot(mean_ice_cover, aes(x = year4, y = mean)) +
  geom_line() +
  labs(x = "Year", y = "Mean number of frozen days", title = "Yearly average duration of ice at Madison Lake Area from 1851 - 2019")

mean_yearly_ice
```

The figure above shows a gradual decline in average duration of ice at the three Madison Area Lakes observed.

### Average air temps at Madison Lake Area

```{r}
daily_temps <- weather %>% 
  filter(month %in% c(12, 1, 2)) %>%
  group_by(year4) %>% 
  summarize(avg_temp_c = mean(ave_air_temp_adjusted))

daily_temps_plot <- ggplot(daily_temps, aes(x = year4, y = avg_temp_c)) +
  geom_line() +
  labs(x = "Year", y = "Average air tmeperature (ºC)", title = "Avergae yearly air temperatures at Madison Lake Area from 1851 - 2019")

daily_temps_plot
```

The above figure shows a possible gradual increase in average air temperature.


```{r}
ice_temps <- full_join(daily_temps, mean_ice_cover, key = year4)
```

### Ice vs temperatures

```{r}
ice_temps_plot <- ggplot(ice_temps, aes(x = avg_temp_c, y = mean)) +
  geom_point() +
  labs(x = "Average air temp (ºC)", y = "Average ice duration")

ice_temps_plot
```

This figure shows a negative correlation between mean air temperatures and mean ice duration.


### Thaw date exploration

```{r}
thaw_date <- ice_cover %>% 
  filter(year4 >= 1970) %>% 
  mutate(ice_off = lubridate::ymd(ice_off)) %>% 
  mutate(day = lubridate::yday(ice_off))
#group_by(year4) %>% 
#summarize(day = mean(day, na.rm = TRUE))

thaw_date_plot <- ggplot(thaw_date, aes(x = year4, y = day)) +
  geom_point(aes(color = lakeid))

thaw_date_plot
```

